---
title: "Analysis"
author: "Bryce DiNardo"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rFIA)
library(dplyr)
library(sf)
library(terra)
library(ggplot2)
install.packages("ranger")
library(ranger)
set.seed(123)
```

Get our data 
we have seedling_sf already need to load in 2022 climate data
```{r}
tmmn_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Tmmn/tmmn_2022.tif")
tmmx_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Tmmx/tmmx_2022.tif")
tmean_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/tmean/tmean_2022.tif")
pr_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Pr/pr_2022.tif")
```

Pull climate data from the pixel in which each FIA plot is located, building training data frame
```{r}
# Extract values from each raster
seedling_climate <- seedling_sf %>%
  mutate(
    tmmn = terra::extract(tmmn_2022, seedling_sf)[,2],
    tmmx = terra::extract(tmmx_2022, seedling_sf)[,2],
    tmean = terra::extract(tmean_2022, seedling_sf)[,2],
    pr = terra::extract(pr_2022, seedling_sf)[,2]
  )
#Drop geometry col, dont need it
training_df <- seedling_climate %>%
  st_drop_geometry()
#need to make owner stuff into factors
training_df$OWNCD <- as.factor(training_df$OWNCD)
training_df$OWNGRPCD <- as.factor(training_df$OWNGRPCD)
str(training_df)
```
training_df is ready to run through the ranger function

##Running RF
Standard
```{r}
rf_model <- ranger(
  formula = TPA ~ ELEV + OWNCD + OWNGRPCD + tmmn + tmmx + tmean + pr,
  data = training_df,
  num.trees = 1000,        # number of trees (default is 500)
  mtry = 3,               # number of variables tried at each split
  importance = "permutation",# or "permutation" for more robust importance
  num.threads = 4,        # adjust to your CPU cores
  seed = 123              # reproducibility
)
rf_model 
rf_model$variable.importance

sqrt(rf_model$prediction.error)

```

Let chanfe importance 
```{r}

barplot(rf_model$variable.importance,
        main = "Permutation Importance",
        las = 2, col = "forestgreen")
#Still sucks but we will add more predictors
```
Elevation most impornat land owner least but is there an issue wth them being catigorical yet as numbers
```{r}
summary(training_df$TPA)
hist(log1p(training_df$TPA))

ggplot(training_df, aes(x = TPA)) +
  geom_histogram(fill = "forestgreen", color = "white", bins = 30) +
  scale_x_log10() +
  labs(
    title = "Histogram of Douglas-fir Seedling Density (Log Scale)",
    x = "Trees per Acre (log10 scale)",
    y = "Number of Plots"
  ) +
  theme_minimal()

```
```{r}
max(training_df$TPA)
```

```{r}
hist(training_df$TPA, breaks = 50, main = "Raw TPA Distribution")
hist(log1p(training_df$TPA), breaks = 50, main = "Log-Transformed TPA Distribution")

```
```{r}
#log transforming greatly improves the models fit
rf_model_log <- ranger(
  formula = log1p(TPA) ~ ELEV + OWNCD + OWNGRPCD + tmmn + tmmx + tmean + pr,
  data = training_df,
  num.trees = 1000,
  mtry = 3,
  importance = "permutation",
  seed = 123
)
rf_model_log
rf_model_log$variable.importance
sqrt(rf_model_log$prediction.error)

#need to compare the orginal model to log transform in the orginal models space
preds_log_back <- exp(rf_model_log$predictions) - 1
obs <- training_df$TPA

# RMSE
rmse_log_back <- sqrt(mean((obs - preds_log_back)^2))

# R²
r2_log_back <- 1 - sum((obs - preds_log_back)^2) / sum((obs - mean(obs))^2)

c(RMSE_raw = sqrt(rf_raw$prediction.error),
  R2_raw = rf_raw$r.squared,
  RMSE_log_back = rmse_log_back,
  R2_log_back = r2_log_back)
```
Graph side by side
```{r}
par(mfrow = c(1,2))

plot(training_df$TPA, rf_raw$predictions,
     xlab = "Observed TPA", ylab = "Predicted TPA",
     main = "Raw Model", pch = 16, col = "forestgreen")
abline(0,1,col="red",lty=2)

plot(training_df$TPA, exp(rf_model_log$predictions)-1,
     xlab = "Observed TPA", ylab = "Predicted TPA",
     main = "Log Model (Back-transformed)", pch = 16, col = "steelblue")
abline(0,1,col="red",lty=2)

```

chageg