---
title: "Stats"
author: "Bryce DiNardo"
date: "2025-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Hypothesis testing

H1: Tmean and Pr will be the most important climatice drivers of seedling desnity

```{r cars}
varimp_long_sum <- models_summary %>%
  unnest(varimp) %>%
  group_by(species, year, predictor) %>%
  summarise(importance = mean(importance), .groups = "drop")

```
How often tmean and pr are top two predictors

```{r pressure, echo=FALSE}
rank_summary <- varimp_long_sum %>%
  group_by(species, year) %>%
  arrange(desc(importance)) %>%
  mutate(rank = row_number())

# Frequency table
rank_summary %>%
  filter(species == "Douglas_Fir") %>% 
  group_by(predictor, species) %>%
  summarise(top2_freq = sum(rank <= 2),
            total_years = n(),
            proportion = top2_freq / total_years) %>% 
  arrange(desc(top2_freq))

```
Something I dont understand yet
```{r}
install.packages("boot")
library(boot)

# Function for bootstrapping mean importance
boot_mean <- function(x, i) {
  mean(x[i])
}


ci_summary <- varimp_long_sum %>%
  group_by(predictor) %>%
  summarise(
    mean_imp = mean(importance),
    ci_low = boot::boot.ci(boot(data = importance, statistic = boot_mean, R = 1000),
                           type = "perc")$percent[4],
    ci_high = boot::boot.ci(boot(data = importance, statistic = boot_mean, R = 1000),
                            type = "perc")$percent[5]
  )

```
Freidman test
```{r}
library(rstatix)

df_imp_long <- varimp_long_sum %>%
  filter(species == "Lodge_Pole") %>%
  select(year, predictor, importance) %>%
  mutate(id = year)   # use year as the blocking factor
friedman_test(df_imp_long, importance ~ predictor | id)
pairwise_wilcox_test(df_imp_long, importance ~ predictor, paired = TRUE, p.adjust.method = "BH")


```
### Addressing H2
Importance of climate varibles are greatest in years with climate anmoarlies

Step 1: Use my gridmet data so determine if there are climate abnormoaloes
using tmean and pr
```{r}
#load in saved gridmet data 
get_tmean<-function(year){
  #make the file path that can be used in interations
tmean_year<-paste0("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/tmean/tmean_", year, ".tif")

#read in the corresponding rasters 
rast(tmean_year)
}


get_pr<-function(year){
  #make the file path that can be used in interations

pr_year<-paste0("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Pr/pr_", year, ".tif")

#read in the corresponding rasters 
rast(pr_year)
}
```
get for 2014
```{r}
tmean_2014<-get_tmean(2014)
pr_2014<-get_pr(2014)

```
Extract means for the whole raster
```{r}
pr_mean_2014 <- global(pr_2014, "mean", na.rm = TRUE)
tmean_mean_2014 <- global(tmean_2014, "mean", na.rm = TRUE)

```
See if I can build a climate summarize table with years and values in rows and tmean and pr as col
```{r}
years <- 2014:2022

# Initialize an empty list to collect rows
rows <- list()

for (y in years) {
  pr_rast <- get_pr(y)
  tmean_rast <- get_tmean(y)
  
  # Extract statewide mean values as numerics
  pr_mean <- global(pr_rast, "mean", na.rm = TRUE)[1]
  tmean_mean <- global(tmean_rast, "mean", na.rm = TRUE)[1]
  
  # Store as a list row
  rows[[as.character(y)]] <- data.frame(
    year = y,
    pr = pr_mean,
    tmean = tmean_mean
  )
}

# Bind all rows together
climate_summary <- do.call(rbind, rows)
years <- 2014:2022

# Initialize an empty list to collect rows
rows <- list()

for (y in years) {
  pr_rast <- get_pr(y)
  tmean_rast <- get_tmean(y)
  
  # Extract statewide mean values as numerics
  pr_mean <- global(pr_rast, "mean", na.rm = TRUE)[1]
  tmean_mean <- global(tmean_rast, "mean", na.rm = TRUE)[1]
  
  # Store as a list row
  rows[[as.character(y)]] <- data.frame(
    year = y,
    pr = pr_mean,
    tmean = tmean_mean
  )
}

# Bind all rows together
climate_summary <- do.call(rbind, rows)
rownames(climate_summary) <- NULL
names(climate_summary)[2:3] <- c("pr", "tmean")
names(climate_summary)[2:3] <- c("pr", "tmean")
climate_summary


```
Calcute how geeat  each years given pr and tmean is from the standard devioiation is from the overall mean, is it is greater than 1 in either direction it is considered an extereme climate year 
```{r}
threshold <- 1  

climate_summary<-climate_summary %>%
    mutate(
      pr_z = (pr - mean(pr, na.rm = TRUE)) / sd(pr, na.rm = TRUE),
      tmean_z = (tmean - mean(tmean, na.rm = TRUE)) / sd(tmean, na.rm = TRUE),
      category = case_when(
        abs(pr_z) > threshold | abs(tmean_z) > threshold ~ "Extreme",
        TRUE ~ "Average"))
climate_summary
```
Now each year is in a caregory of either an extrerem or average climate year, so I can join it with my permutaion imporantace table I made from all my random forests
```{r}
varimp_with_climate <- varimp_long %>%
  left_join(climate_summary %>% select(year, category), by = "year")

```
other stuff
```{r}
names(varimp_with_climate)

climate_importance <- varimp_with_climate %>%
  filter(species == "Douglas_Fir", predictor %in% c("pr", "tmean", "tmmn", "tmmx")) %>%
  group_by(year, category) %>%
  summarise(total_importance = sum(importance), .groups = "drop")

```

stat
```{r}
wilcox.test(total_importance ~ category, data = climate_importance)

```
vis
```{r}
ggplot(climate_importance, aes(x = category, y = total_importance, fill = category)) +
  geom_boxplot() +
  labs(title = "Climate predictor importance in Extreme vs. Average years",
       y = "Total climate importance", x = "Year category")
`
```


