---
title: "Get_Seedling"
author: "Bryce DiNardo"
date: "2025-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rFIA)
library(dplyr)
library(sf)
library(terra)
library(ggplot2)
```

## Getting and Processing FIA Seedling desnsity data  

###Load FIA tables and clip to most recent inventory years

```{r}
or_fia <- readFIA("C:/Users/Bryce/OneDrive - University Of Oregon/FIA", nCores = 4)
or_fia_clip <- clipFIA(or_fia, mostRecent = TRUE)
```
### Function to get seedling data from your specific invenotry year
Inputs: year of intrest and the speceis code for speceis of intrest found on FIA appendix F
Output: returns a data table with plot cn, sub plot number, condition id, inveotry year, species code, seedling count used in calulations, trees per arcer (TPA) unadjusted to plot.
```{r}
get_seedling<-function(year, spcd){
  Seedling_df<-or_fia_clip$SEEDLING %>% 
  filter(SPCD == spcd, INVYR == year, !is.na(TREECOUNT_CALC), TREECOUNT_CALC > 0) %>% 
  select(PLT_CN, SUBP, CONDID, INVYR, SPCD, TREECOUNT_CALC, TPA_UNADJ)
  return(Seedling_df)
}
```
Calcuate trees per an acer per plt_cn. In puts
```{r}
seedling_TPA <- function(get_seedling_df){
  get_seedling_df %>% 
    group_by(PLT_CN) %>% 
  summarise(TPA = mean (TPA_UNADJ, na.rm = TRUE))
}
```
Plot level data
```{r}
get_lat_long<-function(year){
  or_fia_clip$PLOT %>%
  filter(INVYR == year) %>% 
  select(PLT_CN, LAT, LON, ELEV) %>%
  filter(!is.na(LAT), !is.na(LON), !is.na(ELEV))
}
```
Get owner information
```{r}
get_owncd<-function(year){
  or_fia$COND %>% 
  filter(INVYR == year) %>% 
  select(PLT_CN,OWNCD, OWNGRPCD)
}
```
Time to join them all together
```{r}
join<-function(tpa_df, lat_long_df, owner_df){
  tpa_df_latlong<-left_join(tpa_df, lat_long_df, by = "PLT_CN") %>% filter(!is.na(TPA), TPA > 0)
  left_join(tpa_df_latlong, owner_df,by = "PLT_CN")
}
```
Go for df to a spacital feature
```{r}
seeding_as_sf<-function(joined_df){
  seedling_sf <- st_as_sf(joined_df, coords = c("LON", "LAT"), crs = 4326)
  return(seedling_sf)
}

```
##Combined all functions 
```{r}
process_seedlings <- function(year, spcd, save_path = NULL) {
  
  # 1. Get seedling records
  Seedling_df <- or_fia_clip$SEEDLING %>%
    filter(SPCD == spcd,
           INVYR == year,
           !is.na(TREECOUNT_CALC)) %>%
    select(PLT_CN, SUBP, CONDID, INVYR, SPCD, TREECOUNT_CALC, TPA_UNADJ)
  
  # 2. Calculate plot-level TPA (sum of TPA_UNADJ across subplots)
  # Note: FIA defines TPA_UNADJ as already count × expansion factor
  Seedling_tpa <- Seedling_df %>%
    group_by(PLT_CN) %>%
    summarise(TPA = sum(TPA_UNADJ, na.rm = TRUE), .groups = "drop")
  
  # 3. Get plot coordinates and elevation
  lat_long_df <- or_fia_clip$PLOT %>%
    filter(INVYR == year) %>%
    select(PLT_CN, LAT, LON, ELEV) %>%
    filter(!is.na(LAT), !is.na(LON), !is.na(ELEV))
  
  # 4. Get ownership info
  owner_df <- or_fia$COND %>%
    filter(INVYR == year) %>%
    select(PLT_CN, OWNCD, OWNGRPCD)
  # 5. Join everything
  joined_df <- Seedling_tpa %>%
    left_join(lat_long_df, by = "PLT_CN") %>%
    filter(!is.na(TPA), TPA > 0) %>%
    left_join(owner_df, by = "PLT_CN")
  
  # 6. Convert to sf object
  seedling_sf <- st_as_sf(joined_df, coords = c("LON", "LAT"), crs = 4326)
  
  # 7. Optionally save to file
  if (!is.null(save_path)) {
    st_write(seedling_sf, save_path, delete_dsn = TRUE)
  }
  
  return(seedling_sf)
}

```
Test
```{r}
process_seedlings(2021, 202)
```



