---
title: "Analysis"
author: "Bryce DiNardo"
date: "2025-11-19"
output: html_document
---
###Used analysis function and loops found at bottom
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rFIA)
library(dplyr)
library(sf)
library(terra)
library(ggplot2)
library(ranger)
set.seed(123)
```

Get our data 
we have seedling_sf already need to load in 2022 climate data
```{r}
tmmn_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Tmmn/tmmn_2022.tif")
tmmx_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Tmmx/tmmx_2022.tif")
tmean_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/tmean/tmean_2022.tif")
pr_2022<-rast("C:/Users/Bryce/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Processed Climate Tiffs/Pr/pr_2022.tif")
```

Pull climate data from the pixel in which each FIA plot is located, building training data frame
```{r}
# Extract values from each raster
seedling_climate <- seedling_sf %>%
  mutate(
    tmmn = terra::extract(tmmn_2022, seedling_sf)[,2],
    tmmx = terra::extract(tmmx_2022, seedling_sf)[,2],
    tmean = terra::extract(tmean_2022, seedling_sf)[,2],
    pr = terra::extract(pr_2022, seedling_sf)[,2]
  )
#Drop geometry col, dont need it
training_df <- seedling_climate %>%
  st_drop_geometry()
#need to make owner stuff into factors
training_df$OWNCD <- as.factor(training_df$OWNCD)
training_df$OWNGRPCD <- as.factor(training_df$OWNGRPCD)
str(training_df)
```
training_df is ready to run through the ranger function

##Running RF
Standard
```{r}
rf_model <- ranger(
  formula = TPA ~ ELEV + OWNCD + OWNGRPCD + tmmn + tmmx + tmean + pr,
  data = training_df,
  num.trees = 1000,        # number of trees (default is 500)
  mtry = 3,               # number of variables tried at each split
  importance = "permutation",# or "permutation" for more robust importance
  num.threads = 4,        # adjust to your CPU cores
  seed = 123              # reproducibility
)
rf_model 
rf_model$variable.importance

sqrt(rf_model$prediction.error)

```

Let chanfe importance 
```{r}

barplot(rf_model$variable.importance,
        main = "Permutation Importance",
        las = 2, col = "forestgreen")
#Still sucks but we will add more predictors
```
Elevation most impornat land owner least but is there an issue wth them being catigorical yet as numbers
```{r}
summary(training_df$TPA)
hist(log1p(training_df$TPA))

ggplot(training_df, aes(x = TPA)) +
  geom_histogram(fill = "forestgreen", color = "white", bins = 30) +
  scale_x_log10() +
  labs(
    title = "Histogram of Douglas-fir Seedling Density (Log Scale)",
    x = "Trees per Acre (log10 scale)",
    y = "Number of Plots"
  ) +
  theme_minimal()

```
```{r}
max(training_df$TPA)
```

```{r}
hist(training_df$TPA, breaks = 50, main = "Raw TPA Distribution")
hist(log1p(training_df$TPA), breaks = 50, main = "Log-Transformed TPA Distribution")

```
```{r}
#log transforming greatly improves the models fit
rf_model_log <- ranger(
  formula = log1p(TPA) ~ ELEV + OWNCD + OWNGRPCD + tmmn + tmmx + tmean + pr,
  data = training_df,
  num.trees = 1000,
  mtry = 3,
  importance = "permutation",
  seed = 123
)
rf_model_log
rf_model_log$variable.importance
sqrt(rf_model_log$prediction.error)

#need to compare the orginal model to log transform in the orginal models space
preds_log_back <- exp(rf_model_log$predictions) - 1
obs <- training_df$TPA

# RMSE
rmse_log_back <- sqrt(mean((obs - preds_log_back)^2))

# R²
r2_log_back <- 1 - sum((obs - preds_log_back)^2) / sum((obs - mean(obs))^2)

c(RMSE_raw = sqrt(rf_raw$prediction.error),
  R2_raw = rf_raw$r.squared,
  RMSE_log_back = rmse_log_back,
  R2_log_back = r2_log_back)
```
Graph side by side
```{r}
par(mfrow = c(1,2))

plot(training_df$TPA, rf_raw$predictions,
     xlab = "Observed TPA", ylab = "Predicted TPA",
     main = "Raw Model", pch = 16, col = "forestgreen")
abline(0,1,col="red",lty=2)

plot(training_df$TPA, exp(rf_model_log$predictions)-1,
     xlab = "Observed TPA", ylab = "Predicted TPA",
     main = "Log Model (Back-transformed)", pch = 16, col = "steelblue")
abline(0,1,col="red",lty=2)

```

```{r}
seedling_sf_2019<- st_read("C:/Users/dinardo/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Extracted_Seedling_Climate/Douglas_Fir/Douglas_Fir_2019.gpkg")
colnames(seedling_sf_2019)
str(seedling_sf_2019)
seedling_sf_2019$OWNCD<-as.factor(seedling_sf_2019$OWNCD)
st_drop_geometry(seedling_sf_2019)

```


###FUNCTION



Function to get extracted .gdpk ready to be fed into ranger
```{r}
get_training_data<-function(year, species_name){
  #Load in the extracted file previously created
   gpkg_file <- file.path(
  "C:/Users/dinardo/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Extracted_Seedling_Climate",
  species_name,
  paste0(species_name, "_", year, ".gpkg")
)
   #remove geometry to make it a data frame
  extracted_file <- st_read(gpkg_file, quiet = TRUE)
  extracted_file<-st_drop_geometry(extracted_file)
  #set owncd and owngrpcd to factor
  extracted_file$OWNCD<-as.factor(extracted_file$OWNCD)
  extracted_file$OWNGRPCD<-as.factor(  extracted_file$OWNGRPCD)
  
#return the ranger ready data.frame 
  return(extracted_file)
}
```
Test and look at data and try different log transforms for best fit
```{r}
df_2019<-get_training_data(2019, "Western_Hemlock")
hist(log10(df_2019$TPA))
hist(df_2019$TPA)
```
playng with running a random forest
```{r}
rf_model_log <- ranger(
  formula = log10(TPA) ~ ELEV + OWNCD + OWNGRPCD + tmmn + tmmx + tmean + pr,
  data = df_2019,
  num.trees = 1000,
  mtry = 3,
  importance = "permutation",
  seed = 123
)
rf_model_log
rf_model_log$variable.importance
```
###FUNCTION for running and saving ranger outputs for multple years/speceis combo
```{r}

fit_and_save_ranger <- function(year, species_name, output_dir = "C:/Users/dinardo/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Ranger_Outputs") {
  # Load training data
  df <- get_training_data(year, species_name)
  
  # Fit ranger model
  rf_model <- ranger(
    formula = log10(TPA) ~ ELEV + OWNCD + OWNGRPCD + tmmn + tmmx + tmean + pr,
    data = df,
    num.trees = 1000,
    mtry = 3,
    importance = "permutation",
    seed = 123
  )
  
  # Create output path
  output_file <- file.path(output_dir, paste0("rf_model_", species_name, "_", year, ".rds"))
  
  # Save model object
  saveRDS(rf_model, output_file)
}
```
Test
```{r}
fit_and_save_ranger(2019, "Douglas_Fir")
```
```{r}
rf_2019<-readRDS("C:/Users/dinardo/OneDrive - University Of Oregon/Data_Sci/Seedling_Density_BD/data/Ranger_Outputs/rf_model_Douglas_Fir_2019.rds")
rf_2019$variable.importance
```
Loop through all years for douglis fir
```{r}
year<-2014:2022

for(i in year){
  fit_and_save_ranger(i, "Douglas_Fir")
}
```
Western Hemlock
```{r}
for(i in year){
  fit_and_save_ranger(i, "Western_Hemlock")
}
```
Lastly Lodge Pole
```{r}
for(i in year){
  fit_and_save_ranger(i, "Lodge_Pole")
}
```

