---
title: "Seedling_processing2.0"
author: "Bryce DiNardo"
date: "2025-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rFIA)
library(dplyr)
library(sf)
library(terra)
library(ggplot2)
```

## I hope this works
Right now just looking to get 2022 Douglas-fir TPA and then join with lat long by plot to make a map 
Get the data tables load up and clip to just the most recent invenotry year set 2012-2022
```{r cars}
or_fia <- readFIA("C:/Users/Bryce/OneDrive - University Of Oregon/FIA", nCores = 4)
or_fia_clip <- clipFIA(or_fia, mostRecent = TRUE)
```
Filter SEEDLING table to get douglas-fir and inventory year 2022
```{r}
head(or_fia_clip$SEEDLING)
Seedling_df<-or_fia_clip$SEEDLING %>% 
  filter(SPCD == 202, INVYR == 2022, !is.na(TREECOUNT_CALC), TREECOUNT_CALC > 0) %>% 
  select(PLT_CN, SUBP, CONDID, INVYR, SPCD, TREECOUNT_CALC, TPA_UNADJ)

head(Seedling_df)

```
To calculate TPA (trees per acre) for each plot, we first tidy the seedling data by filtering for Douglas-fir (SPCD == 202) in the 2022 inventory. Each row represents a subplot (SUBP) within a plot (PLT_CN), and includes an estimate of seedling density (TPA_UNADJ) for that subplot.
Since each plot typically contains four subplots (SUBP 1–4), we sum the TPA_UNADJ values across all subplots that share the same PLT_CN. This gives us the total seedling density (TPA) for Douglas-fir in each plot, expressed in trees per acre.
```{r}
Seedling_tpa<-Seedling_df %>% 
  group_by(PLT_CN) %>% 
  summarise(TPA = sum (TPA_UNADJ, na.rm = TRUE))
```
We see a lot of 74.96528 and 149.93057, that is bacasue the expansiton factor is for every 1 seedling in mircoplot there is 74.96528 seedlings in the acre, so lots of plots of plot have just 1 or 2 seedlings


Next I need to join the plot coordinates to the TPA calculations 
```{r}
or_fia_clip$PLOT
#Get PLT_CN, LAT, LON 
plot_coords <- or_fia_clip$PLOT %>%
  filter(INVYR == 2022) %>% 
  select(PLT_CN, LAT, LON) %>%
  filter(!is.na(LAT), !is.na(LON))
#Join the two tables
seedling_spatial<-left_join(Seedling_tpa, plot_coords, by = "PLT_CN") %>% filter(!is.na(TPA), TPA > 0)

```

